\section{Conclusions and outlook}\label{conclusions}
Rapid and equitable decarbonization of the building heat sector is an indispensable cornerstone in a sustainable society. Special attention is needed for the rented residential buildings sector since a sustainable investment decision is in the landlord's hands. Simultaneously, an expected increase in the CO\textsubscript{2} price primarily impacts the tenant's energy costs. This work studies cost-optimal subsidy payment strategies incentivizing sustainable heat system change and retrofitting measures at the multi-apartment building level. We analyze the results of the application of the developed modeling framework to a partly renovated old building switching either to the district heating network or implementing an air-sourced heat pump system under several decarbonization storylines.\vspace{0.5cm}

We find that a fair and equitable switch to a sustainable heat system is possible but with massive public subsidy payments. In particular, the building's owner (landlord) investment grant and additional rent-related revenues based on the building renovation are crucial to trigger the profitability of the investment. At the same time, subsidy payments to the tenants are required at the beginning of the investment period to limit the energy and rent-related spendings. Furthermore, the results impressively show that the heat pump alternative is not competitive in supplying heat service need in partly renovated old buildings. Either the subsidy payments are significantly higher than in the district heating case, or the equitable constraints of the model can not be satisfied. Deep building renovation and associated reduction of heat demand enable to feasible solutions but with high total costs because passive retrofitting measures need to be incentivized too.\vspace{0.5cm}

Furthermore, the results demonstrate that allocating the costs of inaction (i.e., sticking to the existing gas-based heating system and paying the increasing CO\textsubscript{2} costs) between the governance, the building owner (landlord), and the tenants is an important lever and can reduce the required subsidy payments. First and foremost, the biggest drop of the total subsidies (to nearly half) takes place when the costs of inaction are completely borne by the building owner (landlord). Also, a decrease in the landlord's interest rate reduces the total costs but limits the maximum share of the costs of inaction allocated to the landlord and implies a lower bound of the cost-minimized solution.\vspace{0.5cm}

Future work may investigate a stronger coupling of active and passive renovation measures as a necessary condition for subsidy payments. This could bring further insights to decarbonization strategies with an eye on the heat demand and sustainable heat source alternatives in the multi-apartment residential building sector (i.e., climate neutrality in 2050). Besides, the tenant's diversification within the building could be improved (e.g., different willingness to pay to contribute to CO\textsubscript{2} mitigation). More generally, this study could be extended by introducing further technology options, such as solar photovoltaic, solar thermal, and heat and electricity storage systems. 