\documentclass[review]{elsarticle}
\input{packages.tex}

\begin{document}
\begin{frontmatter}

\title{Equitable decarbonization of heat supply in residential multi-apartment rental buildings: Optimal subsidy allocation between the property owner and tenants}
\author[1]{Sebastian Zwickl-Bernhard\corref{cor1}}
\ead{zwickl@eeg.tuwien.ac.at}
\author[1]{Hans Auer}
\author[1]{Antonia Golab}
\cortext[cor1]{Corresponding author}
\address[1]{Energy Economics Group (EEG), Technische Universität Wien, Gusshausstrasse 25-29/E370-3, 1040 Wien, Austria}

\begin{abstract}
	The core objective of this work is to demonstrate equitable decarbonization of heat supply in residential multi-apartment rental buildings. A modeling framework is developed determining a socially balanced financial governance support strategy between the property owner and tenants to trigger a heating system change. The results of different decarbonization scenarios of a partly renovated old building switching from gas-fired heat supply to either the district heating network or being equipped with a heat pump system show that an equitable heating system change is possible, but with massive public subsidy payments. Particularly, the investment grant to the property owner and additional rent-related revenues due to building renovation are decisive for the profitability of the investment. Simultaneously, subsidy payments to the tenants are required at the beginning of the investment period to limit their energy and rent-related spendings. Results also show that the heat pump alternative is not competitive compared with district heating, even in case of extensive retrofitting of the building. Allocating the costs of inaction (opportunity costs associated with rising CO\textsubscript{2} prices) between the governance, property owner, and tenants turns out as an important lever, as required subsidy payments can be reduced significantly.
\end{abstract}

\begin{keyword}
	Equitability, decarbonization, residential, heat supply, subsidy payments, heat system change, property owner, tenants
	
\end{keyword}
\end{frontmatter}

\section*{Nomenclature}
\begin{center}
	\renewcommand{\arraystretch}{1.1}
	\centering
	\small
	\begin{tabular}{lm{8cm}r}
		Type & Description & Unit\\
		\hline
		Set and index & & \\
		\hline
		
		{$y \in \mathcal{Y}=\{1,\ldots,Y\}$} & Years, index by $y$\\
		{$m \in \mathcal{M}=\{1,\ldots,M\}$} & Months, index by $m$\\
		\hline
		Decision variables\\
		\hline
		{$\Psi$} & Investment grant to the property owner & \SI{}{EUR}\\
		{$\Omega_{y,m}$} & Subsidy payment to a tenant in $y$ and $m$ & \SI{}{EUR}\\
		{$d_{y,m}$} & Total heat demand per tenant/unit in $y$ and $m$& \SI{}{kWh}\\
		{$q_{y,m}$} & Heat demand supplied by the new heating system alternative in $y$ and $m$& \SI{}{kWh}\\
		{$\pi$} & Capacity of the new heating system alternative & \SI{}{kW}\\
		{$r_{y,m}$} & Rent charge adjustment in $y$ and $m$ & \SI{}{EUR\per m^2}\\
		\hline
		Relevant parameters\\
		\hline
		{$n$} & Number of tenants within the multi-apartment building & \SI{}{1}\\
		{$i$} & Interest rate of an agent (governance, property owner, tenant)& \SI{}{\%}\\
		{$d_{y,m}$} & Total heat demand per unit in $y$ and $m$ & \SI{}{kWh}\\
		{$\alpha_{m}$} & Load factor (ratio total and peak demand) in $m$ & \SI{}{1}\\
		{$c_{alt}$} & Investment costs of the heat system alternative& \SI{}{EUR \per kW}\\
		{$c_{con}$} & Construction costs (for adaption of one dwelling/unit) of the heat system alternative per unit & \SI{}{EUR}\\
		{$\bar{r}$} & Initial rent price & \SI{}{EUR \per m^2}\\
		{$\rho$} & Upper limit of the biannual rent charge adjustment & \SI{}{\%}\\
		{$a$} & Rented area per tenant/unit & \SI{}{m^2}\\
		{$p_{init,y}$} & Energy price fueling the initial heating system & \SI{}{EUR \per kWh}\\
		{$p_{alt,y,m}$} & Energy price fueling the heating system alternative & \SI{}{EUR \per kWh}\\
		\hline
	\end{tabular}
\end{center}
\newpage

\input{1_Introduction.tex}
\input{2_Stateoftheart.tex}
\input{3_Methodology.tex}
\input{4_Results.tex}
\input{5_Conclusions.tex}

\section*{Declaration of interests}
None.
\section*{Declaration of Competing Interest}
The authors report no declarations of interest.
\section*{Acknowledgments}
This project has received funding from the European Union's Horizon 2020 Research and Innovation Programme under Grant Agreement No. 835896. The authors acknowledge TU Wien Bibliothek for financial support through its Open Access Funding Programme. The authors would like to thank Christian Kalchschmied and Andreas Lux for their valuable contributions from the field to this work.

\bibliography{mybibfile}
\appendix
\setcounter{table}{0}
\setcounter{figure}{0}
\newpage
\section{Data}\label{app:data}
Table \ref{tab:a1} shows specific emissions, energy prices, and further technical assumptions. The values correspond to the initial input parameters in $2025$ in our analysis. Furthermore, it is assumed that the specific emissions of electricity and district heating decrease linearly between $2025$ and the corresponding decarbonization target year of the scenario ($2040$ in the \textit{Directed Transition} and \textit{Societal Commitment} scenario as well as $2050$ in the \textit{Gradual Development scenario}). The energy price development of electricity, natural gas, and district heating is in line with the assumptions in \cite{fina2019profitability}.  According to this, the (retail) electricity price increases by \SI{2.37}{\%} and the district heating price by \SI{5}{\%} per year. Additionally, the CO\textsubscript{2} price increases the energy price according to the specific emissions per year. Table \ref{tab:time} shows the CO\textsubscript{2} price development in the different scenarios.

\begin{table}[h]
	\centering
	\scalebox{0.85}{
		\renewcommand{\arraystretch}{1.35}
		\begin{tabular}{llrr}
			\toprule 
			Variable & Unit & Value & Ref.\\\hline
			Specific emissions$\vert$Electricity & \SI{}{kgCO_{2} \per kWh} & \SI{0.130}{}&\cite{emissions}\\
			Specific emissions$\vert$District heating & \SI{}{kgCO_{2} \per kWh} & \SI{0.132}{}& \cite{dh_emissions}\\
			Specific emissions$\vert$Natural gas & \SI{}{kgCO_{2} \per kWh} & \SI{0.220}{}&\cite{emissions}\\
			Price$\vert$District heating & \SI{}{EUR \per kWh} & \SI{0.047}{}& \cite{dh_prices}\\
			Price$\vert$Natural gas & \SI{}{EUR \per kWh} & \SI{0.050}{}& \cite{gas_prices}\\
			Price$\vert$Electricity & \SI{}{EUR \per kWh} & \SI{0.200}{}& \cite{electricity_prices}\\
			Coefficient of performance (average) & \SI{}{1} & \SI{2.35}{}& \cite{cop_hp}\\
			\bottomrule
	\end{tabular}}
	\caption{Relevant economic parameters and further empirical settings for Austria in 2020}
	\label{tab:a1}
\end{table}



\begin{table}[h]
	\centering
	\scalebox{0.85}{
		\renewcommand{\arraystretch}{1.35}
		\begin{tabular}{c|c|c|c|c}
			\toprule 
			Scenario ($\SI{}{EUR \per tCO_2}$)& $2020$ & $2025-30$ & $2030-35$ & $2035-40$\\\hline
			\textit{Directed Transition} & $30$ & $196$ & $357$ & $510$\\
			\textit{Societal Commitment} & $30$ & $62$ & $137$ & $273$\\
			\textit{Gradual Development} & $30$ & $83$ & $128$ & $183$\\
			\textit{Low Development}  & $30$ & $60$ & $70$ & $80$\\
			\bottomrule
	\end{tabular}}
	\caption{CO\textsubscript{2} price development}
	\label{tab:time}
\end{table}

\section{Passive building retrofitting measures}\label{app:passive}
We consider passive retrofitting measures in this study in a very simplified way and focus here on the insulation of the building skin and the wall to neighboring buildings only. The economic and technical assumptions are oriented to the study from Fina et al. in \cite{fina2020profitability}. The following relationships between the specific heat demand and the heat pump's (average) coefficient of performance (COP) are assumed: \SI{130}{kWh \per m^2} (COP$=2.5$), \SI{115}{kWh \per m^2} ($3.0$), \SI{100}{kWh \per m^2} ($3.5$). 

\section{Empirical settings of the case example}\label{app:verify}
\begin{table}[h]
	\centering
	\scalebox{0.8}{
		\renewcommand{\arraystretch}{1.35}
		\begin{tabular}{llr}
			\toprule 
			Variable & Unit & Value\\\hline
			Investment cost (heat pump) & \SI{}{EUR \per kW} & \SI{1000}{}\\
			Construction cost & \SI{}{EUR} & \SI{1000}{}\\
			Initial rent price & \SI{}{EUR \per m^2} & \SI{10}{}\\
			Rented area & \SI{}{m^2} & \SI{100}{}\\	
			Total heat demand & \SI{}{kWh} & \SI{22000}{}\\		
			Peak heat demand & \SI{}{kW} & \SI{13}{}\\	
			CO\textsubscript{2} price (2025-2034) & \SI{}{EUR \per tCO_{2}} & \SI{50}{}\\
			CO\textsubscript{2} price (2035-2040) & \SI{}{EUR \per tCO_{2}} & \SI{100}{}\\
			Natural gas price & \SI{}{EUR \per kWh} & \SI{0.05}{}\\
			Electricity price & \SI{}{EUR \per kWh} & \SI{0.2}{}\\
			Specific emissions$\vert$Electricity & \SI{}{kgCO_{2} \per kWh} & \SI{0.130}{}\\
			\bottomrule
	\end{tabular}}
	\caption{Case example's parameters and assumptions}
	\label{tab:a2}
\end{table}

\section{Varying allocation of the costs of inaction}\label{app:varying}
This work considers the CO\textsubscript{2} price-related costs as the costs of inaction and opportunity costs (OC) respectively. Hence, Equation \ref{app:eq:inaction} describes the costs of inaction per year $y$ and month $m$

\begin{align}\label{app:eq:inaction}
	OC_{y,m} =    \gamma_{init} \cdot p_{y}^{CO\textsubscript{2}} \cdot d_{y,m}
\end{align}

where $\gamma_{init}$ is the specific emissions of the initial heating system (i.e., natural gas) and $p_{y}^{CO\textsubscript{2}}$ the CO\textsubscript{2} price in year $y$ and month $m$. Exemplarily, Equation \ref{app:eq:landlord1} shows the property owner's net present value in total when a part of the total OC is allocated to the property owner's net present value

\begin{align}\label{app:eq:landlord1}
	OC_{l} =  \sum_{y} \sum_{m} s_l \cdot \frac{OC_{y,m}}{(1+i_l)^y} 
\end{align}

where $s_l$ is the share in the costs of inaction borne by the property owner. Consequently, Equation \ref{c:npv} is modified as follows by considering the property owner's costs of inaction.

\begin{align}\label{app:eq:landlord}
	-OC_{l} =  -\zeta + \sum_{y} \sum_{m} \frac{1}{(1+i_l)^y} \cdot \lambda_{y,m}
\end{align}

A similar logic is developed in the modification of the tenant's net present value. The tenant's share of the costs of inaction ($OC_{t}$) are considered in Equation \ref{c:ten4}. The tenant's OCs influence the initial spendings that are assumed to be the limit in the sustainable heating system alternative (see Equation \ref{c:app:tenant}).

\begin{align}\label{c:app:tenant}
	K_{alt} = K_{init} - OC_{t}
\end{align}
\end{document}